%=============================================================================
% CONCLUSION
%=============================================================================
\section{Conclusion}
\label{sec:conclusion}

We presented \textsc{Holograph}, a sheaf-theoretic framework for LLM-guided causal discovery.
By formalizing local causal beliefs as presheaf sections and global consistency
as descent conditions, we provide principled foundations for integrating
LLM knowledge into structure learning.

Our key contributions include:
\begin{itemize}
    \item The Algebraic Latent Projection for handling hidden confounders
    \item Natural gradient descent with Tikhonov regularization for optimization
    \item EFE-based active query selection for efficient LLM utilization
    \item Comprehensive sheaf axiom verification revealing fundamental locality failures
\end{itemize}

The systematic failure of the Locality axiom is perhaps our most significant finding.
It demonstrates that the presheaf of ADMGs does not form a classical sheaf
when latent variables induce non-local coupling. This provides a formal measure
of the ``non-sheafness'' inherent in causal models with hidden confounders---a
quantity that could guide future algorithms in detecting latent variable necessity.

\paragraph{Limitations.}
\begin{itemize}
    \item \textbf{Scalability:} Performance on graphs with $n > 100$ variables degrades
          due to $O(n^3)$ projection costs. Sparse approximations may help.
    \item \textbf{LLM Reliability:} Current approach assumes LLM responses are locally consistent.
          Adversarially contradictory LLMs could violate this assumption.
    \item \textbf{Identifiability:} As with all causal discovery methods, we can only recover
          structure up to Markov equivalence without interventional data.
\end{itemize}

\paragraph{Future Work.}
Promising directions include:
\begin{enumerate}
    \item \textbf{Cohomological Measures:} Develop sheaf cohomology metrics to quantify
          Locality violations, potentially using $\check{\text{C}}$ech cohomology.
    \item \textbf{Hybrid Methods:} Combine \textsc{Holograph} with constraint-based algorithms
          (e.g., FCI) to leverage both continuous optimization and discrete constraint propagation.
    \item \textbf{Interventional Extensions:} Extend to experimental design settings where
          interventions can be performed, potentially enabling full causal identification.
\end{enumerate}

\paragraph{Speculative Connections.}
We note a suggestive parallel between our Locality failure and quantum non-locality.
In quantum mechanics, entangled systems violate Bell inequalities through correlations
that resist local hidden variable explanations. Similarly, ADMGs with latent confounders
exhibit correlations between distant variables that cannot be explained by local restrictions.
The scaling $\text{Error} \propto \sqrt{n}$ in both settings hints at deeper mathematical
connections---a direction for future theoretical exploration.
